Holomorphic field theories and higher algebra

### 2024-09-30 draft.tex
hopf

Associative algebras arise in quantum mechanics, we have argued, because the spacetime of a mechanical system is a line (``time'').
If we organize the observables (or operators) by when they happen in time,
it corresponds to writing them left-to-right as products in an algebra.
A natural question to ask then is what kind of higher algebraic structure captures how observables behave for theories in higher dimensions?
The shortest answer is ``factorization algebras,'' and we will describe and explore this notion in Section~\ref{sec fact alg},
but there are nice answers for two special classes of theories that we will develop now:
\begin{itemize}
\item For a topological field theory on $\RR^d$ (which we'll define in a moment),
the observables form an algebra over the {\em little $d$-discs operad}~$\Disc_d$ (sometimes also known as the $E_d$ operad). 
Equivalently, one says the observables are an $E_d$-algebra. 
This notion was introduced by algebraic topologists in the 1960s, 
so there is a rich body of work to draw upon.
\item For a holomorphic field theory on $\CC^d$,
the observables form an algebra over the $d$-polydiscs operad~$\PDisc_d$.
This notion was introduced only recently in~\cite{CG1}.
\end{itemize}
We will make the assertions above precise (in part by adding some natural but necessary hypotheses on a theory)
and develop them in parallel.
As a particularly illuminating example, we will explore the $d=1$ case rather closely and explain in what sense a vertex algebra is a kind of ``holomorphic associative algebra.''
We will also discuss the polydiscs algebras for which the higher Kac-Moody algebras are a Lie-algebraic shadow.

@@

@@

We are again thinking about configuration spaces --- here of polydiscs inside larger polydiscs --- but we are remembering the complex structure explicitly.
To describe observables, we want to produce a dg colored operad that encodes the holomorphic nature of this situation.
We do {\em not} want to take singular chains, which would only remember the homotopy type;
instead, we choose to take the Dolbeault complex of each configuration space.
Note, however, that taking the Dolbeault complex is a contravariant functor,
so we get a {\em co}\/operad rather than an operad.
This notion bears the same relationship to operad as a coalgebra does to an algebra;
one simply reverses all the arrows and asks for an appropriate variant of the relations.
For instance, there is a composition
\begin{multline}
\label{pdcirc}
\circ_i: \Omega^{0,\bu}(\PDisc_d(r_1, \ldots, s_1, \ldots, s_m, r_{i+1}, \ldots, r_n| R)) \to \\
\Omega^{0,\bu}(\PDisc_d(r_1, \ldots, r_n| R)) \otimes \Omega^{0,\bu}(\PDisc_d(s_1, \ldots, s_m| S))
\end{multline}
\owen{I don't know how to eliminate the two numbers}
\brian{I like using multline instead of align}
when $r_i = S$, and in that case, the composition $\mu \circ_i \nu$ is given by embedding an output polydisc into an input polydisc of the same radius.
(Composition is not defined if the radii do not match.)

@@

\brian{Hopf surfaces}

>>>>

Associative algebras arise in quantum mechanics, we have argued, because the spacetime of a mechanical system is a line (``time'').\footnote{In the context of holomorphic field theory on $\C^n$, we have treated the radial direction as the ``time''.}
If we organize the observables (or operators) by when they happen in time,
it corresponds to writing them left-to-right as products in an algebra.
A natural question to ask then is what kind of higher algebraic structure captures how observables behave for theories in higher dimensions?
The shortest answer is ``factorization algebras,'' and we will describe and explore this notion in \S~\ref{sec fact alg},
but there are nice answers for two special classes of theories that we will develop now:
\begin{itemize}
\item For a topological field theory on $\RR^d$ (which we'll define in a moment),
the observables form an algebra over the {\em little $d$-discs operad}~$\Disc_d$ (sometimes also known as the $E_d$ operad). 
Equivalently, one says the observables are an $E_d$-algebra. 
This notion was introduced by algebraic topologists in the 1960s, 
so there is a rich body of work to draw upon.
\item For a holomorphic field theory on $\CC^d$,
the observables form an algebra over the $d$-polydiscs operad~$\PDisc_d$.
This notion was introduced only recently in~\cite{CG1}.
\end{itemize}
We will make the assertions above precise (in part by adding some natural but necessary hypotheses on a theory)
and develop them in parallel.
As a particularly illuminating example, we will explore the $d=1$ case rather closely and explain in what sense a vertex algebra is a kind of ``holomorphic associative algebra.''
We will also discuss the polydiscs algebras for which the higher Kac-Moody algebras are a Lie-algebraic shadow.

@@

There is an explicit model for an algebra over the cohomology of the $E_d$-operad called a $P_d$-algebra.
Essentially, these are graded commutative algebras with a Poisson bracket of cohomological degree $-d+1$.
In \cite{CostelloScheimbauer} a parallel model for the cohomology of the holomorphic $\PDisc_d$ colored operad was proposed; though the exact axioms were not written down.
In recent work, similar models to the one given in \cite{CostelloScheimbauer} have appeared using so-called (multivariable) `$\lambda$-brackets' \cite{BGKWY,GKW}. This work uses such an algebraic structure to give a beautiful proposal for computing Feynman diagrams in holomorphic (and mixed holomorphic-topological) field theories (see the next section for more on renormalization within the context of holomorphic QFT).
Such models are similar in spirit to Lie conformal algebras which use the notion of single-variable $\lambda$-bracket to encode the `singular part' of the data of a vertex algebra.

@@

We are again thinking about configuration spaces --- here of polydiscs inside larger polydiscs --- but we are remembering the complex structure explicitly.
To describe observables, we want to produce a dg colored operad that encodes the holomorphic nature of this situation.
We do {\em not} want to take singular chains, which would only remember the homotopy type;
instead, we choose to take the Dolbeault complex of each configuration space.
Note, however, that taking the Dolbeault complex is a contravariant functor,
so we get a {\em co}\/operad rather than an operad.
This notion bears the same relationship to operad as a coalgebra does to an algebra;
one simply reverses all the arrows and asks for an appropriate variant of the relations.
For instance, there is a composition
\begin{multline}
\label{pdcirc}
\circ_i: \Omega^{0,\bu}(\PDisc_d(r_1, \ldots, s_1, \ldots, s_m, r_{i+1}, \ldots, r_n| R)) \to \\
\Omega^{0,\bu}(\PDisc_d(r_1, \ldots, r_n| R)) \otimes \Omega^{0,\bu}(\PDisc_d(s_1, \ldots, s_m| S))
\end{multline}
when $r_i = S$, and in that case, the composition $\mu \circ_i \nu$ is given by embedding an output polydisc into an input polydisc of the same radius.
(Composition is not defined if the radii do not match.)

@@

For complex manifolds, and holomorphic factorization algebras on them, we consider an explicit example that generalizes a well-known concept in complex dimension one.
First, consider a \textit{real} one-dimensional \textit{topological} factorization algebra $\cA$.
If we identify $\cA$ with the associative algebra that it determines, then the factorization homology $\int_{S^1} \cA$ along the circle agrees with the Hochschild cohomology of the associative algebra $\cA$.
In other words, it is the home for all of the `traces' which the algebra $\cA$ admits.
It is well-known that the elliptic curve, or genus one, conformal blocks of a vertex algebra $\mathbb{V}$ is the natural `home' for the characters of the vertex algebra.
The most basic character of a conformal vertex algebra is the $q$-character defined in terms of a trace:
\beqn
\op{char}_{\mathbb{V}} (q) = \op{Tr}_{\mathbb{V}} (q^{L_0-c/24}) ,
\eeqn
where $L_0$ denotes the action of the vector field $z \del_z$.
The shift by the central charge guarantees that for $\mathbb{V}$ conformal that this expression has nice modularity properties.
From the point of view of factorization algebras, we can understand this trace in terms of Hochschild homology of the \textit{modes} algebra associated to $\mathbb{V}$ as introduced in \S\ref{??}.
Indeed, suppose we know that $\mathbb{V}$ corresponds to a holomorphic factorization algebra $\cF_{\mathbb{V}}$ on $\C$.
Then, if $r \colon \C^\times \to \RR_{>0}$ is the radial projection, we have discussed how to extract the modes algebra from the one-dimensional factorization algebra $r_* \cF_{\mathbb{F}}$.
If we further put this on the circle, we find Hochschild homology.
Thus, for every modulus $q$ one has
\beqn
\int_{E_q} \cF_{\mathbb{V}} = \int_{S^1} r_* \cF_{\mathbb{V}} = \op{Hoch}(r_* \cF_{\mathbb{V}}) .
\eeqn
For more on this perspective see \cite{GuiLi}.
From this point of view we hope it is clear how the factorization homology along an elliptic curve encodes characters in the sense of vertex algebras.

In higher dimensions we can build a similar picture.
The simplest so-called \textit{Hopf manifolds} are complex manifolds of the form
\beqn
H_{\mathbf{q}} = \left(\C^d - \{0\}\right) \slash \left((z_1,\ldots,z_d) \sim (q_1 z_1,\ldots,q_d z_d) \right)
\eeqn
where $\mathbf{q} = (q_1,\ldots,q_d)$ are complex numbers satisfying $0 < |q_i| < 1$.
For generic $\mathbf{q}$ such a complex manifold is diffeomorphic to $S^{2d-1} \times S^1$.
The factorization homology of a holomorphic factorization algebra $\cF$ along a Hopf surface of this kind can be identified with the Hochschild homology of the associative algebra of $S^{2n-1}$-modes:
\beqn
\int_{H_{\mathbf{q}}} \cF \simeq \op{Hoch}(r_* \cF)
\eeqn
where, again, $r\colon \C^d - \{0\} \to \RR_{>0}$ is the radius.
Thus, we see that the factorization homology $\int_{H_{\mathbf{q}}} \cF$ naturally encodes traces of the $S^{2n-1}$-mode algebra associated to $\cF$.
In \cite{SWchar}, we have shown how an important invariant called the \textit{supersymmetric index} (or supersymmetric partition function) of a supersymmetric theory along a sphere $S^{2n-1}$ be cast into this framework at the level the holomorphic twist.
